---
title: "Assignment 2"
author: "Ella Brayshaw"
date: "2023-03-13"
output: html_document
---


### Introduction
In this report I will be describing and evaluating the BOP2 trial design. I will begin by explaining the key features and steps. Then I will apply the BOP2 method to the reserve trial scenario that we have been provided with. 

We are interested in looking at the progression between stage 1 to stage 2, then from stage 2 to a definitive phase III trial; these progressions are based on the posterior probability of futility. To begin to show the meaning of this idea, I will set $y_i$ to be the number of observations in the $n_i$ particpants ($n_1$ = number of participants in stage 1 and $n_2 = n_1 + \tilde{n_2}$ = participants at the end of stage 2 ). $y_i$ follows a binomial distribution with probability of success, $\theta$, following a  prior beta distribution with parameters $a_0$ and $b_0$.

After observing the response $y_i$ we can then calculate the posterior distribution using Bayes Theorem :  
$$ P(\theta | y_i) = \frac{P(y_i|\theta)*P(\theta)}{P(y_i)} $$
 This results in the posterior probability also following a beta distribution : $P(\theta|y_i)$ ~ beta($a_i$, $b_i$), where $a_i = a + y_i$ and $b_i = b + n_i - y_i$. 
 
We use this posterior probability to determine progression through the probability of futility. We define the probability of futility as the probability of a response, $\theta < 0.5$ given the observed responses : 
$$P(\theta < 0.5 | y_i) $$
We ideally want this probability to be low as the probability of observing a response is poor. To decide whether we carry on to the next stage we need to define a threshold which varies at each stage to compare this too, we define this threshold to be $$C(n_i) = 1 - \lambda(\frac{n_i}{n_2})^{\gamma}$$ In this report we will investigate choices of these stopping rule parameters $\lambda$ and $\gamma$
I defined functions which calculated the probability of futility as well as the thresholds in the Functions.R folder ......


I will now apply the BOP2 trial design to the RESERVE trial. The RESERVE (pREciSion markERs for VEnetoclax in multiple myeloma) trial is a clinical trial which is looking at the effects of treating patients with relapsed/refractory multiple myeloma (RRMM) with Vendex in combination with CyVenDex. The RESERVE trial is currently in phase II and we are wanting to investigate if the treatment has a large enough impact on tumor size to warrant a larger phase III trial. As there is uncertainty to the effectiveness and toxicity of the drug we want to work in 2 stages where the first stage has a lower sample size. This has better ethical, financial and efficiency considerations. In the this report i will be using $n_1$ participants in stage one of the trial and $\tilde{n_2}$ participants in stage 2, this is if we observe enough responses in stage 1 to move onto stage 2. This leads to a total of $n_2 = n_1 + \tilde{n_2}$ participants at the end of stage 2. 

To begin my exploration of the RESERVE trial I wan to find the optimal values of $\lambda$ and $\gamma$ for a fixed $n_1$ and $n_2$ which result in  a type I error rate of at most 0.05 under the null hypothesis of $H_0 : \theta =0.5$ and the type II error rate of at most 0.2 under the alternative hypothesis of $H_1 : \theta = 0.7$. Therefore I will evaluate the type I and type II error rates over a grid of values for $\lambda$ and $\gamma$ and observe which of these have the results we want. For the bounds for the parameters I used the fact that $0\leq \lambda \leq1$ and then a plot to choose $\gamma \ge0$ :


```{r, echo=FALSE, fig.show="hold", out.width="50%"}

g<-seq(0,10,length.out = 100)

plot(1-0.3*(0.5)^g, ylab = "Decision threshold", main = "Plot for lambda = 0.3")
abline(v=60, col="magenta")

plot(1-0.7*(0.5)^g, ylab = "Decision threshold", main = "Plot for lambda = 0.7")
abline(v=60, col="magenta")
```

Here I have looked at how the decision threshold, $C(n_i) = 1 - \lambda(\frac{n_i}{n_2})^{\gamma}$, changes for different values of $\gamma$ ( which I created with a sequence ). I chose the value of $\frac{n_i}{n_2}=0.5$, as $n_1 < n_2$ hence the faction will always be less than 1. These plots show that beyond a vale of $\gamma = 6$ there is little change to the value of the decision threshold, $C(n_i)$. Hence I will chose the bound $0\le \gamma \le 6$. I have the created a grid of both the values of $\lambda$ and $\gamma$ over the chosen intervals and have named this "lambda_gamma": 

```{r}
lambda_gamma_seq <- expand.grid(lambda = seq(0, 1, length.out = 50),
                       gamma = seq(0, 6, length.out = 50))
```

I have chosen the length of each vector for the parameter to be 50 as ....

To evaluate the type I and type II error rate over this grid I require to call on a function that I have coded; such functions I have stored in a R script of this R project called "Function.R". Using the command : 

```{r}
source("./R/functions.R")
```

I can easily make use of these functions without having to include the code each time. 

In here is a function called "Type1_exact", which I will now use to evaluate the probability of a type 1 error over the values in "lambda_gamma". A type 1 error is such that we declare a successful trial at the end of stage 2, under the null hypothesis that $\theta = 0.5$. This means that we require $P(\theta < 0.5 | y_1) < C(n_1)$ and $P(\theta < 0.5 | y_1, y_2) < C(n_2)$, we require stage 1 to be successful to move to stage 2 hence both of these are needed. To begin I will set $n_1 = 30$ and $n_2 = 60$. 



```{r}
source("./R/functions.R")
T1s <- apply(lambda_gamma_seq, 1, evaluate_seq_T1, n1=30, n2=60)
```

The code above shows me applying the grid of $\lambda$'s and $\gamma$'s over a function named "evaluate sequence" which allows us to evaluate another function named "Type1_exact" over the grid. The Type1_exact is a function of $\lambda$, $\gamma$, $n_1$ and $n_2$ which also uses another function in the folder, this function is called "prob_y1_y2" and it calculates the probability of observing the responses $y_1$ and $y_2$ within the number of participants $n_1$ and $n_2$ respectively. 

To see which values of $\lambda$ and $\gamma$ are required to get a Type I error of $\le 0.05$, I will plot Type 1 error vs $\lambda$ and $\gamma$ separately. the plot for $\gamma$ vs Type I error here appears to be very unhelpful, hence i will begin by choosing a optimal $\lambda$. The clustering effect shown in graph is given by each cluster representing a different value of the other parameter $\gamma$, I have added lines (blue at Type 1 = 0.05 and green at $\lambda = 0.96$), I observe that to the right of $\lambda = 0.96$, the Type I error is always $\le 0.05$. However for certain choices of $\gamma$, the combination with $\lambda = 0.94$ and $\lambda = 0.92$ results in the desired bound for type I error.

```{r, echo = FALSE, fig.show="hold", out.width="50%"}
seq_T1s_bind <- cbind(lambda_gamma_seq, T1s)
seq_T1s_bind$lambda <- round(seq_T1s_bind$lambda, 2)
seq_T1s_bind$gamma <- round(seq_T1s_bind$gamma, 2)


plot(seq_T1s_bind$lambda,seq_T1s_bind$T1s)
abline(h=0.05, col = "blue")
abline(v=0.96, col ="green")

plot(seq_T1s_bind$gamma,seq_T1s_bind$T1s)
```

Currently thr plot of $\gamma$ against type I error is fairly useless, so to look further into optimal values for $\gamma$, I will re-create the plot for only the suitable values of $\lambda$ mentioned above.  

```{r, echo = FALSE, fig.show="hold", out.width="50%"}
sub1 <- seq_T1s_bind[seq_T1s_bind$lambda == 0.98, ]  
sub2 <- seq_T1s_bind[seq_T1s_bind$lambda == 0.96, ]
sub3 <- seq_T1s_bind[seq_T1s_bind$lambda == 0.94, ]
sub4 <- seq_T1s_bind[seq_T1s_bind$lambda == 0.92, ]

plot(sub3$gamma, sub3$T1s)
abline(h=0.05, col = "red")
plot(sub4$gamma, sub4$T1s)
abline(h=0.05, col = "red")
```
From these plots I observe that for $\lambda \in [0.92,0.94]$, to achieve a Type 1 error $\le 0.05$ we would require $\gamma = 0$ which is not a reasonable choice as $\gamma$ is restricted to $[0,\infty]$. Hence I I conclude that for to achieve the maximum value of 0.05 for type I error I will choose $\lambda \in [0.96,1]$ and to find an optimal $\gamma > 0$ I will move on to exploring the type II error rate. 

I will now look at the optimal values for the type II error rate; I begin by evaluating a function named "evaluate_seq_T2" (which lives in the Functions.R folder) over my previous grid of $\lambda$'s and $\gamma$'s and for n1=30 and n2=60 again. A type II error rate is defined, in this setting, as calling a unsuccessful trial at the end of either stage 1 or stage 2. This would happen if $P(\theta < 0.5 | y_1) > C(n_1)$ (failing to continue at stage 1) or $P(\theta < 0.5 | y_1) < C(n_1)$ and $P(\theta < 0.5 | y_2) < C(n_2)$ (calling a successful trial at stage 1 but not at stage 2). This idea is laid out in a function named "Type2_exact" in the Functions folder, which implements the idea by using "(!go_1)|(go_2)" which in R means "don't go at stage 1 or don't go at stage 2 ". 


```{r}
source("./R/functions.R")
T2s <- apply(lambda_gamma_seq, 1, evaluate_seq_T2, n1=30, n2=60)
```

I the have gone on to plot the Type II error rates vs the decision parameters $\lambda$ and $\gamma$. On the plot of $\lambda$ vs type II error rates I have added blue vertical lines to represent the interval I decided for $\lambda$ from the type I error, this shows there are several combinations with $\gamma$ which result in the desired range for type II error. 

```{r, echo = FALSE, fig.show="hold", out.width="50%"}
seq_T2s_bind <- cbind(lambda_gamma_seq, T2s)
#Make values of lambda and gamma 2 decimal places 
seq_T2s_bind$lambda <- round(seq_T2s_bind$lambda, 2)
seq_T2s_bind$gamma <- round(seq_T2s_bind$gamma, 2)

plot(seq_T2s_bind$lambda, seq_T2s_bind$T2s)
abline(h = 0.2, col="red")
abline(v = c(0.96,0.98), col = "blue")

plot(seq_T2s_bind$gamma, seq_T2s_bind$T2s)
abline(h = 0.2, col = "red")
```
By observation of the $\gamma$ plot I see that there appears to a cluster linked to a value of $\lambda$ where type II error always equals 1, hence to find what this value of $\lambda$ is I will make a subsection of results which only includes rows where type 2 error = 1. From the print out of this I see that $\lambda = 1$ results in this type II error, this agrees with my previous conclusion I made so can be disregarded. 

```{r}
seq_T2s_bind$T2s <- round(seq_T2s_bind$T2s, 3)

sub5 <- seq_T2s_bind[seq_T2s_bind$T2s == 1, ]
head(sub5)
# so when lambda = 1 T2 not in the range

```

Finally, to pick my value for $\gamma$ I will make 2 further subsections of the results, one for when $\lambda = 0.98$ and one for $\lambda = 0.96$, as these match with my previous interval I have decided for $\lambda$.

```{r, echo = FALSE, fig.show="hold", out.width="50%"}
# Type 1 errors lowest for lambda = 0.98
sub6 <- seq_T2s_bind[seq_T2s_bind$lambda == 0.98, ]
plot(sub6$gamma, sub6$T2s)
abline(h = 0.2, col = "red")

sub7 <- seq_T2s_bind[seq_T2s_bind$lambda == 0.96, ]
plot(sub7$gamma, sub7$T2s)
abline(h = 0.2, col = "red")
# type 2 error gets smaller as lambda gets smaller
```

From inspection of the plots and the output of the two subsections, I see that the value of type II error drops below 0.2 for $\gamma \ge 0.12$ hence any value of $\gamma \ge 0.12$ would give us the desired results. However we ideally want the probability of type II error to be as low as possible as this means our study would have greater power, this is important as it means we are more likely to detect a true result. To archive this minimal type 11 error I would choose $\lambda = 0.96$ as here there is a minimum type II error of 0.063 versus 0.103 for $\lambda = 0.98$. The plot shows us that the type II error rate appears to be quite consistent after $\gamma =1$ and from the output of the subsection I see that values of $\gamma \ge 2.08$ give the minimum type II error rate of 0.063. I will pick my optimal value of $\gamma$ to be 2.08 for ease as if i wanted to do further investigations it reduces the interval my grid would have to search over. 

```{r, echo = T, results = 'hide'}
T1_lambda0.96 <- seq_T1s_bind[seq_T1s_bind$lambda == 0.96, ]
T2_lambda0.96 <- seq_T2s_bind[seq_T2s_bind$lambda == 0.96, ]
T1_lambda0.96
T2_lambda0.96
```


Hence for the sample size parameters of $n_1 = 30$ and $n_2 = 60$, I have found that the optimal values of $\lambda = 0.96$ and $\gamma = 2.08$. And from inspection of subsections of grids made where $\lambda = 0.96$ I see that these would results a type I error rate of 0.04621577 and type II error rate of 0.063 which satisfy the bounds given in the brief. 
 
I will now go on to investigate how these decsion rule parameters change as the sample size changes, I will follow similar steps to what I carried out previously. To start I shall simply double the previous sample sizes to $n_1 = 60$ and $n_2 = 120$, keeping the 2:1 ratio. So to begin I evaluate the function that calculate probability of Type I and Type II errors but will change the parameter values to the new ones. 

```{r}
source("./R/functions.R")
T1s_60 <- apply(lambda_gamma_seq, 1, evaluate_seq_T1, n1=60, n2=120)
T2s_60 <- apply(lambda_gamma_seq, 1, evaluate_seq_T2, n1=60, n2=120)
```

Making the grids of lambda gammas then type 1 and type 2 

```{r}
seq_T1s_60_bind <- cbind(lambda_gamma_seq, T1s_60)
seq_T1s_60_bind$lambda <- round(seq_T1s_60_bind$lambda, 2)
seq_T1s_60_bind$gamma <- round(seq_T1s_60_bind$gamma, 2)

seq_T2s_60_bind <- cbind(lambda_gamma_seq, T2s_60)
seq_T2s_60_bind$lambda <- round(seq_T2s_60_bind$lambda, 2)
seq_T2s_60_bind$gamma <- round(seq_T2s_60_bind$gamma, 2)

```

plots for type 1 failry simialr to before, same bounds for lambda again 

```{r echo = FALSE, fig.show="hold", out.width="50%"}
plot(seq_T1s_60_bind$lambda,seq_T1s_60_bind$T1s)
abline(h=0.05, col = "blue")

plot(seq_T1s_60_bind$gamma,seq_T1s_60_bind$T1s)
```
some for lambda = 0.94 fall below max 

```{r}
sub8 <- seq_T1s_60_bind[seq_T1s_60_bind$lambda == 0.98, ]  
sub9 <- seq_T1s_60_bind[seq_T1s_60_bind$lambda == 0.96, ]
sub10 <- seq_T1s_60_bind[seq_T1s_60_bind$lambda == 0.94, ]
sub11 <- seq_T1s_60_bind[seq_T1s_60_bind$lambda == 0.92, ]

plot(sub10$gamma, sub10$T1s)
abline(h=0.05, col = "red")
plot(sub11$gamma, sub11$T1s)
abline(h=0.05, col = "red")
```
plots for type 2, points much closer together

```{r}
plot(seq_T2s_60_bind$lambda, seq_T2s_60_bind$T2s_60)
abline(h=0.2, col = "red")
plot(seq_T2s_60_bind$gamma, seq_T2s_60_bind$T2s_60)
abline(h=0.2, col = "red")
```
T2 = 1 when lamba = 1 again so same values
```{r}
seq_T2s_60_bind$T2s_60 <- round(seq_T2s_60_bind$T2s_60, 3)

sub12 <- seq_T2s_60_bind[seq_T2s_60_bind$T2s_60 == 1, ]
head(sub12)
```
```{r}
sub13 <- seq_T2s_60_bind[seq_T2s_60_bind$lambda == 0.98, ]
plot(sub13$gamma, sub13$T2s_60)
abline(h = 0.2, col = "red")

sub14 <- seq_T2s_60_bind[seq_T2s_60_bind$lambda == 0.96, ]
plot(sub14$gamma, sub14$T2s_60)
abline(h = 0.2, col = "red")

sub15 <- seq_T2s_60_bind[seq_T2s_60_bind$lambda == 0.94, ]
plot(sub14$gamma, sub14$T2s_60)
abline(h = 0.2, col = "red")
```
for type 2 all except when lambda = 1 are below range, lambda = 0.94 leads to lowest type 2, so need gamma which leads to tyupe 1 below range. this shows gamma to be in (0, 0.24]. take of 0.24 as gets lower t2 but little change to t1. Gets T1 error of 0.045 and type 2 of 0.011
```{r}
sub16 <- seq_T1s_60_bind[seq_T1s_60_bind$lambda == 0.94, ]
head(sub16)

head(sub15)
```





